The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-10 kalman_walk_paper.tex
more edits on overleaf

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

\usepackage{multicol}

@@

\begin{document}

@@

What are the fitness consequences?
A population of $F_2$s will begin to be substantially less fit than the parentals
once they differ from the optimum by a distance of order $\gamma$,
i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
This implies that hybrid incompatibility among $F_2$s should appear on a time scale of 
$N_e \gamma / \sigma_N$ generations.
The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig},
this deviation of the mean from optimum grows with the square of the distance between the parental populations,
and so we expect fitness costs in $F_1$s to appear on a time scale of $N_e^2$ generations.

If a population has a Gaussian distribution of trait values with variance $\sigma^2$
and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
then the population mean fitness is 
\begin{align*}
   \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
   &=
   \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
\end{align*}
If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
(i.e., genetic load) are small,
but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
(i.e., $\gamma/\sigma$).
This implies that we expect $\sigma_S < \gamma$
but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
in $F_2$ hybrids.
Said another way,
if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
\begin{align*}
   \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
\end{align*}

The fitness consequences of system drift above were due to increased \emph{variance} in $F_2$ hybrids.
If the optimal set $\allS$ is curved (which it will be in general),
then the mean $F_1$ offspring will fall away from the optimum slightly,
leading to an additional fitness cost in hybrids.
However,
as shown in appendix \ref{apx:away_from_opt} and depicted in figure \ref{fig:osc_incompat},
the deviation of the mean from the optimum grows proportionally to $T/N_e$.

\begin{align*}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} \\
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align*}
With $\sigma/\gamma = c_\sigma \sqrt{T/N_e}$
and $\mu/\gamma = c_\mu T/N_e$,
this is (check!!)
$$
\sqrt{\frac{1}{1 + c_\sigma^2 T/N_e}} \exp\left\{-c_\mu^2 \frac{T^2}{N_e^2} \left(\frac{1}{1 + c_\sigma^2 T/N_e}\right)\right\} \\
$$

%\subsection*{Parameter estimates for real systems}
%\js{I think we should move this section?}
%To translate these results into real predictions, we need to know
%the strength of stabilizing selection on the phenotype,
%and the amount (and structure) of heritable variation in the genotype.
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

@@

\paragraph{Speciation rates under neutrality}
%For cases where system structures and $\gamma$ can be inferred,
%specific speciation rates can be estimated.
%For other situations we offer a back of the envelope calculation.
%
%Consider a population that is distributed on $\mathcal{X}$ such that the magnitude of phenotypically
%divergent $\sigma_S$ and neutral $\sigma_N$ diversity are constrained by selection;
%$\sigma_S$ directly, such that $\sigma_S = \gamma$, and $\sigma_N$ indirectly via recombination load.
%Two reproductively isolated populations will, on average, be a distance of $2 \sigma_N \sqrt{T/N_e}$.
%If we assume that $\sigma_N$ is constrained by a function of $\sigma_S$,
%neutral movement along $\mathcal{X}$ pushes hybrids off of the manifold by $\sigma_S$ per $\sigma_N$,
%Lastly, if fitness is Gaussian: $\fit_2 = \exp\left(- d(x, x')^2/\beta^2\right)$,
%and we assume that the strength of selection is only several multiples of the segregating divergent diversity,
%such that $\sigma_S = c \beta$, we can simplify the expression estimating $F_2$ hybrid fitness drops
%to be,
%
%
%$F_2$s between allopatric populations will suffer an average fitness drop relative to their parents of,
%  \begin{align}
%    \exp\left(-\left( \frac{2\sigma_{N}}{\beta}\sqrt{\frac{T}{N_e}} \right)^{2} \right) .
%  \end{align}

%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
%  \begin{align}
%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
%  \end{align}
%  where $\xi := 2\xi_2/\xi_1$.
  Assume that perpendicular variation segregating within a population is constrained by selection such that $\sigma_S/\gamma = \xi$, and that neutral variation is on the order of perpendicular variation, such that $\sigma_N = \sigma_S$. Since the typical distance between two individuals drawn from separate allopatric populations will be $z = 2 \sigma_N \sqrt{T/N_e}$, we expect a drop in hybrid fitness of
  \begin{align}
  \fit_2/\fit_0 = \exp \left(- 4\xi^2 \frac{T}{N_e} \right) .
  \end{align}
  Hybrid fitness depression will be significant when $z > \gamma$. 
  If $\xi$ is in $[\sfrac{1}{5},1]$, speciation will occur on the order of $N_e$ generations, and possibly much faster. Note, however, that if intrapopulation variation is very small such that $\sigma_S \ll \gamma$, speciation rates will be slower.

If $\sigma_N \approx \sigma_S \approx \gamma$ and variance in $F_2$s increases linearly with distance betw parents,
then fitness starts to drop at $T/4N_e$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gaussian load}

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm, xfrac}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\usepackage{lineno}
\usepackage{multicol}

@@

\begin{document}
\linenumbers

@@

What are the fitness consequences?
A population of $F_2$s will begin to be substantially less fit than the parentals
once they differ from the optimum by a distance of order $\gamma$,
i.e., once $\sqrt{4 \omega T/N_e} \approx \gamma / \sigma_N$.
This implies that hybrid incompatibility among $F_2$s should appear on a time scale of 
$N_e (\gamma / \sigma_N)^2 / (4 \omega)$ generations.
The $F_1$s will not suffer fitness consequences until the hybrid mean is further than $\gamma$ from the optimum;
as shown in appendix \ref{apx:away_from_opt} (and suggested by figure \ref{fig:conceptual_fig}),
this deviation of the mean from optimum grows with the square of the distance between the parental populations,
and so we expect fitness costs in $F_1$s to appear on a time scale of $N_e^2$ generations.

% If a population has a Gaussian distribution of trait values with variance $\sigma^2$
% and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
% then the population mean fitness is 
% \begin{align*}
%    \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
%    &=
%    \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
% \end{align*}
% If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
% (i.e., genetic load) are small,
% but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
% (i.e., $\gamma/\sigma$).
% This implies that we expect $\sigma_S < \gamma$
% but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
% in $F_2$ hybrids.
% Said another way,
% if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
% then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
% \begin{align*}
%    \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
% \end{align*}

For a more concrete prediction, suppose that the distribution among hybrids is Gaussian.
A population whose trait distribution is Gaussian with mean $\mu$ and variance $\sigma$,
has mean fitness 
\begin{align} \label{eqn:gauss_fit}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} .
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align}
This assumes a single trait, for simplicity; the multivariate case is done in appendix \ref{apx:gauss_load}.
A population of $F_2$s will have, as above, variance $\sigma^2 = \sigma^2_S + 4 \omega \sigma^2_N T/N_e$.
The mean diverges with the square of the distance between the parentals
with a speed depending on the local geometry of the optimal set
(calculated in appendix \ref{apx:H_calc}), so we set $\mu = c_\mu \gamma T/N_e$.
The mean fitness in parental populations is as in equation \ref{eqn:gauss_fit} with $\mu=0$ and $\sigma = \sigma_S$.
This implies that if we define $\fit_2(T)$ to be the mean relative fitness among $F_2$ hybrids
between two populations separated by $T$ generations, 
(i.e., the mean fitness divided by the mean fitness of the parents)
then
\begin{align} \label{eqn:fit2}
\fit_2(T) = 
  \left(1 + \frac{4 \omega (\sigma_N/\gamma)^2}{1 + (\sigma_S/\gamma)^2} \frac{T}{N_e} \right)^{-1/2} 
     \exp\left\{-\left(c_\mu \frac{T}{N_e}\right)^2 \left(\frac{1}{1 + (\sigma_S/\gamma)^2 + 4 \omega (\sigma_N/\gamma)^2 T/N_e}\right)\right\} .
\end{align}
We discuss the implications of this expression in the next section.

\plr{note: should be a plot of the components of this against $T/N_e$ later}

%%%%%%%%%%
\paragraph{Speciation rates under neutrality}
\plr{continue editing here}
(say something about here we interpret the big math expression in words)
From equation \eqref{eqn:fit2} we can observe that
time is always scaled in units of $N_e$ generations,
the population standard deviations are always scaled by $\gamma$,
and the most important term is the rate of accumulation of segregation variance,
$4 \omega (\sigma_N/\gamma)^2$.
All else being equal, this process will lead to speciation more quickly in smaller populations
and in populations with more neutral genetic variation (larger $\sigma_N$).
These parameters are related -- larger populations generally have more genetic variation --
but since these details depend on the situation, we leave these separate.

argue sometimes $\sigma_N \approx \sigma_S \approx \gamma$, especially if $N_e$ is big
(``large species'')

and that sometimes $\sigma_N \approx \sigma_S \ll \gamma$, but only if $N_e$ is small
(``isolated population'')

and that sometimes maaybe $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load
(``species complex'')

but not likely to be other combinations; make plot of \eqref{eqn:fit2} against $T/N_e$ (or even $T$?) in these three situations.

\subsection*{Parameter estimates for real systems}
%\js{I think we should move this section?}
%To translate these results into real predictions, we need to know
%the strength of stabilizing selection on the phenotype,
%and the amount (and structure) of heritable variation in the genotype.
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gaussian load}
\label{apx:gauss_load}
